Plotting simulated PCR results

### 2022-03-24 publication/reviewer_response.Rmd

# Reviewer 1

> This research will be very useful for people working on metabarcoding projects across many fields. The RPS10 loci appears to be well suited for barcoding oomycetes using Miseq sequencing technology and the creation of a website for the community to contribute reference sequences is a good contribution that will give ongoing value to the research community. I have made some suggestions for minor revisions below. As an aside, the website isn't under review here but could I also suggest that you might include some bioinformatic workflows there as well. This would useful for reproducibility for others in the future.

Thank you for your review and constructive feedback!
We are considering ways to improve the website and would like to have a few more tools there in the future.

> 55 : “Evaluated” might be a better word than “Identified” here

Done.

> 57 : Change “amplify” to “amplify this region from”

Done.

> 79: Change “often” to “can”

Done.

> 96: “Eucalypt” should be changed to “Eucalyptus”

Done.

> 109: Change “fungi and bacteria” to “fungi or bacteria”

Done.

> 110: Could you please expand a little about what you mean about activity of oomycetes.

The statement was referring to some types of oomycetes that must be baited during certain seasons when they are sporulating.
However, since that does not apply to all oomycetes and is vague without more explanation, we have removed that portion of the sentence.

> 176: Change “amplification” to “PCR”

Done.

> 185 “Efficacy” might read better than “usefulness”

Done.

> 185 – 188 : the way this is phrased it sound like you are comparing rps10 primers ability to amplify the ITS1 region with the Sapkota primers which they don’t amplify. Could you re word it please to be a clearer.

Done.

> 205: It would be useful and informative if you included the alignment as a supplementary file

!!!!TODO!!!! (Waiting on Frank to get back to us)

> 211: “All known oomycetes” implies you checked every oomycete ever described

Good point. We meant that the goal was for the primers to amplify all oomycetes, but I see how that could be misinterpreted.
We have modified the text.

> 214: Which cutoffs did you use for the parameters such as GC content, hairpins etc when selecting primers?

Rigid cutoffs were not used, but rather people experienced with primer design compared the OligoAnalyzer results to their previous experiences.

> 239: How was the DNA quantified?

We used a Qubit. This information has been added.

> 321 – 322: This first sentence is not required because the information is repeated in more detail below. Please just add the references for DADA2 and Cutadapt where approcpriate below instead.

Done.

> 478 – 494: Were there any oomycetes amplified/identified with the ITS1 primer set that were not identified with the rps10 primers?

Unfortunately the data from the environmental samples cannot be used to answer this question confidently for two reasons:

* The content of the reference databases differs for the two loci, so even if both methods amplified the same organism, the closest match in the reference database of each locus could be to different taxa. Since many ASVs in the environmental samples have no close match in either database (only 16% of oomycete ASVs from environmental samples had a match >95% identical), this error is likely to occur often.
* The limited taxonomic resolution of ITS1, and rps10 to a lesser extent, means that even if both databases have a reference sequence perfectly matching an ASV, there could be other taxa in that database with the same sequence, so which taxon name gets assigned to the ASV is random.

This is why the mock community was useful; in that case we knew all the organisms sequenced had a perfect match in both databases.

> 516 Please change “are generally more diverged from each other” to “are generally more divergent”

Done.

> 540 - 541: Please change “showed” to “predicted” also please change “non-target taxa” to “all non-target taxa tested”

Thanks for pointing that out. Done.

# Reviewer 2

> This study describes the design of a valuable new barcoding locus/assay developed from a database of mtDNA sequence data. Its utility was then compared carefully against an ITS rDNA locus. The methods were used to metabarcode a mixed mock community and on eDNA samples from natural environments. The paper is a well-written account of a valuable breakthrough in oomycete barcoding and should be published.  A key benefit of the rps10 locus compared to the ITS loci primers seems to me the ability to amplify products beyond the Phytophthora and downy mildew genera to include other oomycetes.  I do, however have a few comments suggested changes that I feel would add to value of the publication, in particular in relation to the read ratio in the mock community.  I also feel it may be helpful to detail which of the reverse primers has specificity to which group of oomycetes. Could this be used to give future researchers an option to pick and choose the combination of rps10 primers to focus on the group of oomycetes of particular interest?

> As an aside: note that the primers ITS6 and 7 were never designed to be, or described as, an oomycete-specific assay by Cooke et al (2000) but were simply modifications of the existing White et al (1990) fungal ITS primers to allow more robust amplification of oomycete samples in the laboratory. They were taken up and tested by other authors as ‘oomycete specific’. It was not entirely surprising they were not effective as the discriminating bases are often not at that the 3’ end that would have offered greater specificity.  I say this not to criticise the rps10 assay which looks very powerful, but the point could be made that an ITS-based assay that was as carefully designed as the rps10 one could also offer value for oomycete barcoding.  The assay of Sapkota and Nicolaisen (2015) was not ‘designed’ but tested ‘in hope’ and is maybe not a fair comparison. Having said that, there were no better oomycete ITS assays to be tested so this papers detailed investigation is valuable and will hopefully prevent others using these ITS primers as oomycete-specific in the future. While a comparison may have been with the Scibetta et al (2012) ITS primers which are highly specific to known downy mildew and Phytophthora I understand that these do not amplify other oomycetes and would not have been equivalent. The key advantages of rps10 – oomycete specificity and improved taxonomic resolution - are clear.

These are excellent points.
We have added a sentence to explain the history of ITS6 and ITS7.

> What happens with Nothophytophthora? I could not see them on Figure 2. There is no reason to suggest they would not amplify but perhaps they should be mentioned somewhere?

The data used to make Figure 2 was collected during the initial design of the primers.
At the time Nothophytophthora had just been described and we were not aware of any publicly available rps10 sequences for Nothophytophthora that included the primer binding sites.
As far as we can tell, the rps10 sequences for Nothophytophthora on NCBI currently available do not have the primer binding sites so we cannot predict whether or not Nothophytophthora would be amplified by this method.
Nothophytophthora is one of many genera not represented by currently available rsp10 sequences.
We have added a sentence to the methods (around line 228) using Nothophytophthora as an example for genera we dont have data for.

> The web-site https://nam04.safelinks.protection.outlook.com/?url=http%3A%2F%2Fwww.oomycetedb.org%2F&amp;data=04%7C01%7Cniklaus.grunwald%40oregonstate.edu%7C04b24f72622c4cf9996308da05c111ce%7Cce6d05e13c5e4d6287a84c4a2713c113%7C0%7C0%7C637828624036753211%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C3000&amp;sdata=%2FlQvEfQI7iNfh7VRxfzITmsi4h2dliKEmNs2dGUVj80%3D&amp;reserved=0 is a valuable resource.

Thank you!

> L145 “due to it robust statistical” needs an extra word or ‘its’

Done.

> L153 “currently available ITS1” I would say “some currently available ITS1” as there are ones that do not such as the unmentioned Scibetta et al primers. I feel reference should be made to the source paper (Scibetta et al 2012) which described the downy mildew and Phytophthora-specific primers employed in several key studies such as those of Landa and Riddell.
> Scibetta, S.; Schena, L.; Chimento, A.; Cacciola, S.O.; Cooke, D.E. A molecular method to assess Phytophthora diversity in environmental samples. J. Microbiol. Methods 2012, 88, 356–368, doi:10.1016/j.mimet.2011.12.012.

Good observation.
A reference to the Scibetta paper has been added.

> L157 “as little as 5.3% of” simpler just to say “only 5.3% of ”

Done.

> L185 suggest “power” or “specificity“ rather than ”usefulness” or leave and just say they were compared.

Another reviewer suggested "efficacy" and we used that, but thanks for the suggestion!

> L294 should specify the version of Illumina chemistry employed.  v2 or v3 and how many cycles?  This is important.

Done.

> L471 onwards – for the mock community it would also be interesting to know the read numbers for each species. Were all reads balanced in 1/24 ratio? This would give a measure of the uniformity of amplification efficiency of both assays.  For example did the unknown proportion of downy mildew vs plant DNA result in fewer DM species reads? How did ITS and mtDNA compare? The more uniform amplicon length of rps10 compared to ITS1 should theoretically offer an advantage as it may minimise competition in which shorter amplicons are more abundant than longer ones due to more efficient amplification. Was this the case? I understand the assays are not intended to be quantitative but it would be useful to know if differences in read number broadly align with initial DNA concentration.

Good idea. This information has been added.

> SE Comment: Is there a spot you could include the importance to Phytobiome research, maybe early in the introduction. It is just nice to see the word phytobiome/s used.

Done.

### 2022-03-28 09--simulated_pcr_results.Rmd
added primer binding site analysis requested by reviewer

```{r}
library(ape)
library(metacoder)
library(tibble)
```

## Read in results

@@

## Clean up headers

@@

## Filter out ambigous taxa and atrifacts

@@

## Plot what was amplified

@@

>>>>

```{r}
library(ape)
library(metacoder)
library(tibble)
library(stringr)
library(readr)
library(dplyr)
library(furrr)
library(Biostrings)
```

## Plot Felipe's analysis results

### Read in results

@@

### Clean up headers

@@

### Filter out ambigous taxa and atrifacts

@@

### Plot what was amplified

@@

## Which primers amplify which oomyctes

A reviewer thought it would be useful to know which of the variations of the primer amplified which organisms, with the hope that researchers could use subsets of the primers to target specific groups.

### Extract all oomycete from reference databases

```{r}
rps_ref <- read_fasta(file.path('intermediate_data', 'reference_databases', 'rps10_reference_db.fa'))
rps_ref <- rps_ref[grepl(names(rps_ref), pattern = 'Oomycetes')]
```

Remove spaces from sequences (not sure why there would be spaces).

```{r}
rps_ref <- trimws(gsub(rps_ref, pattern = ' ', replacement = ''))
```

Replace names with just species names

```{r}
names(rps_ref) <- str_match(names(rps_ref), ';([a-zA-Z0-9_\\-\'". ]+);oodb_.+;$')[,2]
```

Replace underscores with spaces

```{r}
names(rps_ref) <- gsub(names(rps_ref), pattern = '_', replacement = ' ')
```

Remove species with ambiguous names or numbers

```{r}
is_ambiguous <- function(n) {
  grepl(n, pattern = ' sp$') |
    grepl(n, pattern = ' spp$') | 
    grepl(n, pattern = ' spp.') |
    grepl(n, pattern = ' sp.') |
    grepl(n, pattern = ' aff. ')  | 
    grepl(n, pattern = ' cf.') |
    grepl(n, pattern = '[0-9]+') 
}
rps_ref <- rps_ref[! is_ambiguous(names(rps_ref))]
```

Remove information below species

```{r}
names(rps_ref) <- sub(names(rps_ref), pattern = ' var\\.?.+$', replacement = '')
names(rps_ref) <- sub(names(rps_ref), pattern = ' subsp\\.?.+$', replacement = '')
```

Now the names should be cleaned up enough:

```{r}
unique(names(rps_ref))
```

### Extract predicted amplicons

Not all reference sequences will have the entire region.
Some will have the region but not the primer sites, so these must be removed.

```{r}
primer_data <- read_csv(file.path("raw_data", "primer_data.csv"))
primer_data <- filter(primer_data, primer_id != 'rps10_f-F', locus == "rps10")
forward <- primer_data$sequence[primer_data$direction == "Forward"]
reverse <- primer_data$sequence[primer_data$direction == "Reverse"]
```

The `matchProbePair` function from the `Biostrings` package can extract a region predicted to be amplified by primers, but for some reason it does not allow for ambiguity codes, even though it is a trivial change to make it do so.
Below is a modified version of the function that does allow ambiguity codes by providing the `fixed` parameter and passing it to the `matchPattern` calls.

```{r}
matchProbePair <- function(Fprobe, Rprobe, subject, algorithm="auto",
             logfile=NULL, verbose=FALSE, fixed = FALSE)
    {
        ## This won't copy the data if Fprobe and Rprobe are already DNAString objects
        F <- DNAString(Fprobe)
        R <- DNAString(Rprobe)

        ## F and R hits on the + strand
        Fp_hits <- start(matchPattern(F, subject, algorithm=algorithm, fixed = fixed))
        Rp_hits <- start(matchPattern(R, subject, algorithm=algorithm, fixed = fixed))

        ## F and R hits on the - strand
        Fm_hits <- end(matchPattern(reverseComplement(F), subject, algorithm=algorithm, fixed = fixed))
        Rm_hits <- end(matchPattern(reverseComplement(R), subject, algorithm=algorithm, fixed = fixed))

        if (verbose) {
            cat("Fp_hits:", Fp_hits, "  Rp_hits:", Rp_hits,
                "  Fm_hits:", Fm_hits, "  Rm_hits:", Rm_hits, "\n")
        }

        matches0 <- Biostrings:::reduceProbePairMatches(c(Fp_hits, Rp_hits), c(Fm_hits, Rm_hits))
        ans <- Views(subject, start=matches0$start, end=matches0$end)

        if (!is.null(logfile)) {
            nFp <- length(Fp_hits)
            nRp <- length(Rp_hits)
            nFm <- length(Fm_hits)
            nRm <- length(Rm_hits)
            nmatches0 <- length(ans)
            ## cat("", ..., sep="\t") is a trick to get an extra tab
            cat("", nFp, nRp, nFm, nRm, nmatches0, file=logfile, sep="\t")
        }
        ans
}
```

The code below extracts predicted amplicons using primer matches, removes the primers, and aligns the trimmed amplicons to the sequences that did not have primer matches to try to find other sequences in the reference databases that do not have primer sequences but are full length matches to those that do.

```{r}
get_amplicon_chr <- function(seq) {
  map2_chr(seq@ranges@start, seq@ranges@width, function(s, w) {
    as.character(seq@subject[seq(from = s, length.out = w)])
  })
}

calc_amplicons <- function(seqs, forward, reverse, min_coverage = 0.9, ...) {
  # Get simulated amplicons using primers
  full_amps <- future_map(seqs, function(s) {
    get_amplicon_chr(matchProbePair(DNAString(s), Fprobe = forward, Rprobe = reverse))
  })

  # Check for multiple possible amplicons per input
  if (any(map_int(full_amps, length) > 1)) {
    stop('Some inputs have more than one amplicons.')
  }
  full_amps <- unlist(full_amps[map_int(full_amps, length) == 1])

  return(full_amps)
}

rps_amps <- calc_amplicons(rps_ref, 
                           forward = primer_data$sequence[primer_data$direction == 'Forward' & primer_data$locus == 'rps10'],
                           reverse = primer_data$sequence[primer_data$direction == 'Reverse' & primer_data$locus == 'rps10'])
```

The amplicon could not be extracted from many sequences, probably because it was incomplete.
Here is the proportion of reference sequences with full length amplicons extracted:

```{r}
length(rps_ref)
length(rps_amps)
```

Lets remove any sequences that are duplicates (same sequence and same species name):

```{r}
rps_amps <- rps_amps[! duplicated(paste0(names(rps_amps), rps_amps))]
length(rps_amps)
```

I noticed when debugging an issue further on in the analysis that at least one sequence is reverse complimented relative to the other sequences.
I will fix that here:

```{r}
to_rev_comp <- ! grepl(rps_amps, pattern = "^GTTGGTT")
rps_amps[to_rev_comp] <- rev_comp(rps_amps[to_rev_comp])
```

### List primer binding site sequences for each species

Cut out the primer binding region for each primer:

```{r}
forward_bindings <- substr(rps_amps, start = 0, stop = nchar(forward))
reverse_bindings <- substr(rps_amps, start = nchar(rps_amps) - nchar(reverse) + 1, stop = nchar(rps_amps)) %>%
  rev_comp()
```

check that the primer binding sites are the right sequences for the primers:

```{r}
stopifnot(all(grepl(forward_bindings, pattern = "^GTTGGTTAGAG[CT]A[AG]AAGACT$")))
stopifnot(all(grepl(reverse_bindings, pattern = "^AT[AG][CT][CT]TAGAAAGA[CT]T[CT]GAACT$")))
```

We can put this information in a table

```{r}
binding_data <- tibble(species = names(forward_bindings),
                       genus = sub(species, pattern = ' .+$', replacement = ''),
                       forward = forward_bindings,
                       reverse = reverse_bindings,
                       amplicon = rps_amps)
```

and save it for later reference:

```{r}
write_csv(binding_data, file = file.path('results', 'primer_binding_site_sequences.csv'))
```

### Check if primers correspond to genus

Some primers are much more common than others:

```{r}
binding_data$primer_pair <- paste(forward_bindings, "+", reverse_bindings)
table(binding_data$forward)
table(binding_data$reverse)
table(binding_data$primer_pair)
```

They don't seems to correspond the genus differences generally, but perhaps a few groups have an unique primer pair:

```{r}
t(table(binding_data[, c("genus", "forward")]))
t(table(binding_data[, c("genus", "reverse")]))
t(table(binding_data[, c("genus", "primer_pair")]))
```
